The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-07-03 paper.tex
Updates from Overleaf

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for deriving different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats and environments.
Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
This study encompasses a multidisciplinary approach, integrating insights from natural, social and cultural sciences.

It's about reading the stories that human remains tell---from sex, age and health to broader aspects like cultural practices and environmental impacts.
Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods.
This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
Sex assessment is a key part of this analysis because it provides essential information about an individual's identity and helps build a biological profile.
This profile offers insights into the person's life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases.

The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts.
It provides a standardized, digital framework for documenting, analyzing, and presenting data from human skeletal remains, making information accessible and usable for future research.
ANNO supports the use of digital 3D models of bones in 3D editors such as AnthroWorks3D, playing a crucial role in categorizing and analyzing data in a flexible yet standardized and machine-readable manner, thus ensuring interoperability across different systems and studies.
Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections \citep{spurensuche}.
Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:

\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
With these remains scheduled for reburial in 2021, digitization became essential.
For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.

\paragraph{Access}\label{sec:access}
Anthropological skeletal material from excavations often remains undetermined due to a shortage of specialists at the respective institutions.
A digital solution would provide greater availability of human skeletal remains and allow for the efficient allocation of anthropologists and collections.
For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists.
For instance, by digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.

\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}. 
For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible.
For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

@@

%\paragraph{Theoretical Background}
% KH: Prepended this paragraph with "Generally" for a slightly better tie-in to the one before. Everyone feel free to improve if you find a better way.
Generally, ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

@@

Tools such as ANNO and AnthroWorks 3D are the driving forces and bridges of digitization, enabling the much-needed digital transformation in anthropological research.
They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage.

\section{State of research}\label{sec:stateofresearch}

\subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
The Anthropological Notation Ontology (ANNO) distinguishes itself by focusing specifically on anthropology rather than general anatomy.
This focus underscores the importance of understanding the relationship between bones and the anatomical and osteometric features found on them.
ANNO emphasizes the need for precise definitions using established anatomical and anthropological terminology, including directional terms, to facilitate clear communication within the field.
This approach allows for a more nuanced understanding of human skeletal structures, essential for anthropological analysis and research.
The Anthropological Notation Ontology (ANNO) aims to map the entirety of anthropological knowledge directly onto a digital 3D skeletal model, functioning as a comprehensive knowledge representation.
It projects anthropological information as a layer onto the skeleton, incorporating and localising aspects and interrelations crucial for anthropological study, effectively serving as a navigable atlas for skeletal analysis.
ANNO is designed to align anthropological knowledge, including measurement explanations and landmark identification, with methods and the skeletal framework itself.
This modeling ensures data compatibility for extensive analysis, integrating a maximal amount of information.
A pivotal advantage of ANNO is its ability to directly associate information with a digital 3D model facilitating the inherent connection to the raw data and enhancing comprehension, reproducibility, and sustainability of documentation.
ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.

This granular method facilitates a deep and scalable anthropological understanding.
ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.

The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances),
which are either sporadically represented or absent in existing ontologies, on the other hand.
ANNO is designed to evolve gradually by integrating more anthropological aspects, ultimately becoming a modular knowledge environment.
This approach allows for step-by-step enhancements to the ontology's scope and utility, enabling a comprehensive integration of anthropological knowledge.

@@

\paragraph{Further research gaps}\label{sec:gaps+}
There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
%
The primary issue lies in the absence of standardization.
While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in practice~\citep{frequencyta,doestamatter,athighlights}.
Instead, the usage of terminology is as fluid as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
Thus, the usage of English designations is preferred in the Anglo-American sphere, while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
Yet, knowledge of Latin is generally declining.
With the terms becoming more abstract to the people using them, there is also a common prevalence for spelling divergences and grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
Meanwhile, the existing terminology is not flexible enough for language-like usage; for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

@@

%\paragraph{Functions}
The functions are divided into discriminant functions for sex determination and regress functions for body height.
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific or similar to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
The exact number of categories may vary depending on the particular method used.
ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
%Names are assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender are noted.
%For the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification are important; for the regression functions, the error interval is important.
%
%While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
%ANNO links to concepts of the FMA when they exist, but structures them in a different way.
%
%\paragraph{Sex determination}
% auto translated from the project report, use as basis:
%
%\paragraph{Regress functions}
%Regress functions for body height and body weight estimation.
%Names were assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
%In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
%
%Standardised description of position
%In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered.
%Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet.
%In the latter case, the OWL file can be generated with a custom OWL generator.

>>>>

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top-level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

Anthropology, as a discipline, explores the full range of human diversity and similarity across cultures, social structures, habitats, and environments.
Anthropological skeletal analysis, a subset of this field, is particularly focused on understanding the human skeletal system.
This study encompasses a multidisciplinary approach, integrating insights from natural, social, ­and cultural sciences.

It's about reading the stories that human remains tell---from sex, age, and health to broader aspects like cultural practices and environmental impacts. Hence, data from historical and prehistoric anthropology~\cite{prehistoricanthropology}, as well as modern forensic science, provide a comprehensive understanding of deceased individuals and past populations from different time periods. This process involves investigating individuals' identity, health status, behavior, lifestyle, culture, and the circumstances of their death~\cite{spurensuche} by determining phenotypes such as sex, age, height, ancestry, and other individualizing traits using anthropological methods.
Sex assessment is a key part of this analysis because it provides essential information about an individual’s identity and helps build a biological profile. This profile offers insights into the person’s life, health, and the population they belonged to, aiding in reconstructing past societies, understanding demographic patterns, and identifying individuals in forensic cases. \\
\\
The Anthropological Notation Ontology (ANNO) is motivated by the need for consistent data recording, enhanced comparability, facilitated access to, and sustainable preservation of human skeletal remains for research and cultural heritage contexts. It provides a standardized, digital framework for documenting, analyzing, and presenting data from human skeletal remains, making information accessible and usable for future research. ANNO supports the use of digital 3D models of bones in 3D editors such as AnthroWorks3D, playing a crucial role in categorizing and analyzing data in a flexible yet standardized and machine-readable manner, thus ensuring interoperability across different systems and studies.
Anthropological analysis of human skeletal remains is inherently comparative, heavily relying on reference collections \citep{spurensuche}. Several key challenges profoundly affect anthropological research \citep{aw3dcidoc}:
\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value. Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial. This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army. With these remains scheduled for reburial in 2021, digitization became essential. For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.
\paragraph{Access}\label{sec:access}
Anthropological skeletal material from excavations often remains undetermined due to a shortage of specialists at the respective institutions. A digital solution would provide greater availability of human skeletal remains and allow for the efficient allocation of anthropologists and collections. For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists. For instance, by digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide. This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.
\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}. A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented. It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}. This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis \cite{HeuschkelSchmiedelLabudde2024}. 
For instance, Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources are compatible. For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.\\
\\

@@

%\paragraph{Theoretical Background}
Generally, ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

@@

Tools such as ANNO and AnthroWorks 3D are the driving forces and bridges of digitization, enabling the much-needed digital transformation in anthropological research. They provide a comprehensive and sustainable solution, enhancing our understanding of human history and cultural heritage

\section{State of research}\label{sec:stateofresearch}
\subsection{Distinguishing features of ANNO}\label{sec:distinguishingAnno}
The Anthropological Notation Ontology (ANNO) distinguishes itself by focusing specifically on anthropology rather than general anatomy.
This focus underscores the importance of understanding the relationship between bones and the anatomical and osteometric features found on them.
ANNO emphasizes the need for precise definitions using established anatomical and anthropological terminology, including directional terms, to facilitate clear communication within the field.
This approach allows for a more nuanced understanding of human skeletal structures, essential for anthropological analysis and research.
The Anthropological Notation Ontology (ANNO) aims to map the entirety of anthropological knowledge directly onto a digital 3D skeletal model, functioning as a comprehensive knowledge representation.
It projects anthropological information as a layer onto the skeleton, incorporating and localising aspects and interrelations crucial for anthropological study, effectively serving as a navigable atlas for skeletal analysis.
ANNO is designed to align anthropological knowledge, including measurement explanations and landmark identification, with methods and the skeletal framework itself.
This modeling ensures data compatibility for extensive analysis, integrating a maximal amount of information.
A pivotal advantage of ANNO is its ability to directly associate information with a digital 3D model facilitating the inherent connection to the raw data and enhancing comprehension, reproducibility, and sustainability of documentation.
ANNO meticulously details skeletal anatomy through an atomic design approach, breaking it down into its most basic, indivisible components, then systematically organizing these components to represent the skeletal structure with high fidelity.
This granular method facilitates a deep and scalable anthropological understanding.
ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.
The current state of ANNO includes a foundational module that focuses on skeletal anatomy on the one hand and anthropological aspects such as osteometric landmarks (measurement points) and measurements (distances), which are either sporadically represented or absent in existing ontologies, on the other hand.
ANNO is designed to evolve gradually by integrating more anthropological aspects, ultimately becoming a modular knowledge environment.
This approach allows for step-by-step enhancements to the ontology's scope and utility, enabling a comprehensive integration of anthropological knowledge.

@@

\paragraph{Further research gaps}\label{sec:gaps+}
There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
%
The primary issue lies in the absence of standardization.
While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
Yet, knowledge of Latin is generally declining.
With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

@@

%\paragraph{Functions}
The functions are divided into discriminant functions for sex determination and regress functions for body height.
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific or similar to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
The exact number of categories may vary depending on the particular method used.
ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
%Names are assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender are noted.
%For the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification are important; for the regression functions, the error interval is important.
%
%While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
%ANNO links to concepts of the FMA when they exist, but structures them in a different way.
%
%\paragraph{Sex determination}
% auto translated from the project report, use as basis:
%
%\paragraph{Regress functions}
%Regress functions for body height and body weight estimation.
%Names were assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
%In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
%
%Standardised description of position
%In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered. Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet. In the latter case, the OWL file can be generated with a custom OWL generator.
